The Strong Gravitational Lens Finding Challenge

### 2019-02-22 paper.tex
few changes by MM

To model the background objects that are lensed we use sources from the Hubble Ultra Deep Field (UDF) that have been decomposed into shapelet functions to remove noise.  This is the same set of images as used in \citet{2008AandA...482..403M,2010AandA...514A..93M}.   There are 9,350 such sources with redshifts and separate shapelet coefficients in 4 bands.

@@

We use {\em Logistic Regression} (LR), a machine learning algorithm,
to train a classifier model.  LR requires a parameter search over the
regression coefficient, $C_{LogReg}$.  The parameters from both the
feature extrator, HOG, and the linear classifier, LR, contain
parameters that we optimize for peak model performance.  We use {\em
  GridSearchCV} from {\em Scikit-learn} to select cross-validated
parameters for HOG parameters and a subset of $C_{LogReg}$ values with
20\% of the test images provided.  We then run a finer parameter
search over $C_{LogReg}$, splitting the test images into 80\% training
and and 20\% test to avoid overfitting the data.  We use the best
parameters to then train the entire dataset for the final model
classifier that we used to evaluate the competition data.

@@

We used a CNN \citep{Fukushima1980,Lecun1998} with a simple architecture of 12 layers (inspired by \citep{symmetry}), see table \ref{tab:architecture}.
To avoid the problem of the data flow distribution getting out of the comfort zone of the activation functions ("Internal Covariate Shift"), we used a mix of normalization propagation \citep{norm_prop} (without the constraint on the weights but a proper initialization) and batch normalization \citep{batch_norm} (slowly disabled over the iterations).
As activation function, we used a scaled and shifted ReLU (Rectifier Linear Unit), 
\begin{equation} \label{eq:relu}
    \frac{1}{\sqrt{\pi-1}} (\sqrt{2 \pi} \max(0,x) - 1),
\end{equation}
to satisfy the properties required by the normalization propagation.
Our batch normalization implementation computes the mean of the activation function $\bar\mu_i$ using the following equation 
\begin{equation} \label{eq:batchnorm}
    \bar\mu_i \longleftarrow (1-\eta) \; \bar\mu_{i-1} + \eta \; \mu_i(\text{batch}).
\end{equation}
$\bar\mu_i$ is computed using the mean value $\mu_i$ over the batch in combination with the previous mean $\bar\mu_{i-1}$ using an inertia value $\eta$ set to $1$ at the beginning and decaying with the iterations.
For the training, the 20,000 provided images were split into two sets, 17,000 for training and 3k for validation.
Each iteration of the gradient descent (more precisely ADAM \citep{adam}) minimizes the cross entropy, 
\begin{equation} \label{eq:xent}
    \left\{
    \begin{array}{ll}
        - \log(p)   & \text{if the image is a true lens} \\
        - \log(1-p) & \text{if the image is a nonlens}
    \end{array}
    \right.,
\end{equation}
where $p$ is the output of the neural network, computed over a batch of 30 images, 15 lenses and 15 nonlenses, picked from the training set.
The small batches with only 30 images were easier to handle computationally but added more noise to the gradient which we considered negligible due to there being only two classes to classify.

@@

GAMOCLASS is a CNN based classifier.
We used the full training data set in the proportion of 4/5 for training and 1/5 for validation. The training images were labelled with 1 if showing strong lensing and 0 otherwise. Our CNN gives as output a probability [0,1] of the input image being a strongly lensed system.
The final architecture of our model is illustrated in Fig. \ref{Fig1_gamoclass}. The input image (101x101 pixels) is first processed by a 2D convolution layer with a 3x3 filter size, then subsampled by a 3x3 max pooling layer.  Another two identical units follow, with a growing dimensionality of the output space %%(depth, i.e. output number of filters)
 in the convolution, for a total of 3 convolutional layers and 3 max pooling layers. Each of these convolutional layers is followed by a rectified linear unit (ReLU) step. The output of these units is then processed through a single fully-connected layer follower by a dropout layer, and, finally, by a one-neuron fully connected layer with sigmoid activation functions. For the classification problem we used the binary cross-entropy cost function and found the weights  using ADAM \citep{Kingma_2014}) optimization method. The use of the ADAM optimizer improved the learning rate compared to tests with stochastic gradient descent (SGD).
In order to increase the size of the training set and make the model invariant to specific transformations, we perform these data augmentation steps:
1) Random rotation of the image in the range [0, 180$\degree$], using a reflection fill mode to keep constant the size of the images,
2) The images were randomly shifted of 0.02 times the total width of the image,
3) Random horizontal and vertical flips of the images.

During the training we initialize the weights of our model with random normal values and we "warm up" the training of the CNN for 25 epochs, using an exponential decay rate ($10^{-6}$) \citep{Huang_2016}  and then a staring learning rate of  0.001. Then the network was trained using an early stopping method, and for a maximum number of 300 epochs. The early stopping method is an effective method of preventing overfitting and consists in stopping the training if a monitored quantity does not improve for a fixed number (called \textit{patience}) of training epochs. The quantity that we monitored was the \textit{accuracy} of the classification of the validation sample. The best architecture was trained over 220 epochs with a parameter of patience equal to 20.
We implemented our code in the Keras framework \citep{Chollet_2015} on top of Theano \citep{Bastien_2012}.
Our architecture converges with a classification accuracy of 91\% on the validation sample. We further evaluated the performance of our classifier calculating the ROC (see section \ref{sec:figure_of_merit})  curve of the classifier, i.e. the True Positive Rate (TPR) against the False Positive Rate. As shown in Fig. \ref{Fig2_gamoclass}, we reached a TPR higher than the 90\% with a FPR < 8\%.

@@

We used a native CNN from Matlab\footnote{\texttt{https://www.mathworks.com/products/matlab.html}, \texttt{https://www.mathworks.com/help/nnet/convolutional-neural-networks.html}}, which has Convolutional 2D layers with 20 5x5 filters. This CNN can work either with one or three input images, representing greyscale and colour images.
We employed different strategies for the two samples available for the challenge, which involve combinations of the available bands running in one or more CNN, using or not the preprocessing, and combining the output with the aid of other machine learning methods.
In each case we used the simulations made available for the challenge both to train and to validate the results and we used the area under the ROC (see section \ref{sec:figure_of_merit}) to determine which combination of methods gives the best result.    
We selected 
%We evaluate the ROC curves for all configurations tested, selecting 
$90\%$ of the images, chosen randomly, for the training and $10\%$ to validate. We repeated the process $10$ times to avoid bias due to a specific choice of training/validation set and to define an uncertainty in our ROC (see section \ref{sec:figure_of_merit}) .

@@

A more detailed description of each configuration tested is presented below:
\begin{enumerate}[label=\Roman*.]
\item Combination of bands r and i with the average between bands u and g. Use one CNN for classification.
\item Creation of $1$ CNN for each band (total of $4$). The $4$ outputs are used as input to a SVM classifier which returns the final classification $p$. 
\item Combination into 4 different combinations of bands:  RGB $\rightarrow$ (\textit{u,g,r}), (\textit{u,g,i}), (\textit{u,r,i}) and (\textit{g,r,i}). One CNN for each combination of bands and then use of the output score as input to an SVM classifier.
%Mediatrix features used as input of an SVM classifier.
%
%IV) Use of wavelet decomposition for each band then use of these decomposed images in the same combinations of case III.
\item  Average of bands in different combinations RGB $\rightarrow$ (\textit{ug,r,i}), (\textit{u,gr,i}) and (\textit{u,g,ri}). The outputs of these 3 CNNs are inputs to a SVM classifier.
\item Use of CNN-activations (CNN feature maps) as inputs to a SVM classifier, using same combinations of bands of III. The output of each CNN is used as input to a SVM classifier.
\item Use of wiener filter and contrast adjustment on each band, then using the resulting images in the same architecture as in (III). 
\end{enumerate}

@@

\subsection{figures of merit}
\label{sec:figure_of_merit}

@@

In addition to these considerations, the performance of a classifier is a function of many characteristics of the lens system.  It might be that one classifier is good at finding systems with large Einstein radii and incomplete arcs, but not as good at finding small complete Einstein rings that are blended with the light of the lens galaxy.  Also a lens may have a source that is too faint to be detected by any algorithm or is too far from the lens to be very distorted, but will be classified as a lens in the test dataset.    We do not impose a definitive arc/ring magnification, brightness or surface brightness limit for a system to be considered a lens because we 
want to include these "barely lensed" objects to test the limits of the classifiers.  As we will see, if you restrict your objectives to detecting only lensed images with surface brightness above some threshold, for example, the "best" algorithm might change and the TPR will change.  For this reason we plot the AUROC, TPR$_0$ and TPR$_{10}$ as a function of several variables for all the entries.  This is done by removing all the lenses that do not exceed the threshold and then recalculating  these quantities, while the number of non-lenses remains the same.

>>>>

To model the background objects that are lensed we use sources from
the Hubble Ultra Deep Field (UDF) that have been decomposed into
shapelet functions to remove noise.  This is the same set of images as
used in \citet{2008AandA...482..403M,2010AandA...514A..93M} \citep[see
also][]{2019MNRAS.482.2823P}.   There are 9,350 such sources with redshifts and separate shapelet coefficients in 4 bands.

@@

We use {\em Logistic Regression} (LR) 
to train a classifier model.  LR requires a parameter search over the
regression coefficient, $C_{LogReg}$.  The parameters from both the
feature extrator, HOG, and the linear classifier, LR, contain
parameters that we optimize for peak model performance.  We use {\em
  GridSearchCV} from {\em Scikit-learn} to select cross-validated
parameters for HOG parameters and a subset of $C_{LogReg}$ values with
20\% of the test images provided.  We then run a finer parameter
search over $C_{LogReg}$, splitting the test images into 80\% training
and and 20\% test to avoid overfitting the data.  We use the best
parameters to then train the entire dataset for the final model
classifier that we used to evaluate the competition data.

@@

We used a CNN \citep{Fukushima1980,Lecun1998} with a simple architecture of 12 layers (inspired by \citep{symmetry}), see table \ref{tab:architecture}.
To avoid the problem of the data flow distribution getting out of the comfort zone of the activation functions ("Internal Covariate Shift"), we used a mix of normalization propagation \citep{norm_prop} (without the constraint on the weights but a proper initialization) and batch normalization \citep{batch_norm} (slowly disabled over the iterations).
As activation function, we used a scaled and shifted ReLU, defined as
\begin{equation} \label{eq:relu}
    \frac{1}{\sqrt{\pi-1}} (\sqrt{2 \pi} \max(0,x) - 1),
\end{equation}
to satisfy the properties required by the normalization propagation.
Our batch normalization implementation computes the mean of the activation function $\bar\mu_i$ using the following equation 
\begin{equation} \label{eq:batchnorm}
    \bar\mu_i \longleftarrow (1-\eta) \; \bar\mu_{i-1} + \eta \; \mu_i(\text{batch}).
\end{equation}
$\bar\mu_i$ is computed using the mean value $\mu_i$ over the batch in combination with the previous mean $\bar\mu_{i-1}$ using an inertia value $\eta$ set to $1$ at the beginning and decaying with the iterations.
For the training, the 20,000 provided images were split into two sets, 17,000 for training and 3k for validation.
Each iteration of the gradient descent (more precisely ADAM \citep{adam}) minimizes the cross entropy, 
\begin{equation} \label{eq:xent}
    \left\{
    \begin{array}{ll}
        - \log(p)   & \text{if the image is a true lens} \\
        - \log(1-p) & \text{if the image is a nonlens}
    \end{array}
    \right.,
\end{equation}
where $p$ is the output of the neural network, computed over a batch of 30 images, 15 lenses and 15 nonlenses, picked from the training set.
The small batches with only 30 images were easier to handle computationally but added more noise to the gradient which we considered negligible due to there being only two classes to classify.

@@

GAMOCLASS is a CNN based classifier.
We used the full training data set in the proportion of 4/5 for training and 1/5 for validation. The training images were labelled with 1 if showing strong lensing and 0 otherwise. Our CNN gives as output a probability [0,1] of the input image being a strongly lensed system.
The final architecture of our model is illustrated in Fig. \ref{Fig1_gamoclass}. The input image (101x101 pixels) is first processed by a 2D convolution layer with a 3x3 filter size, then subsampled by a 3x3 max pooling layer.  Another two identical units follow, with a growing dimensionality of the output space %%(depth, i.e. output number of filters)
 in the convolution, for a total of 3 convolutional layers and 3 max
 pooling layers. Each of these convolutional layers is followed by a ReLU step. The output of these units is then processed through a single fully-connected layer follower by a dropout layer, and, finally, by a one-neuron fully connected layer with sigmoid activation functions. For the classification problem we used the binary cross-entropy cost function and found the weights  using ADAM \citep{Kingma_2014}) optimization method. The use of the ADAM optimizer improved the learning rate compared to tests with stochastic gradient descent (SGD).
In order to increase the size of the training set and make the model invariant to specific transformations, we perform these data augmentation steps:
1) we introduce random rotations of the image in the range [0, 180$\degree$], using a reflection fill mode to keep constant the size of the images;
2) the images are randomly shifted of 0.02 times the total width of the image;
3) the images are randomly flipped horizontally and vertically.

During the training we initialize the weights of our model with random normal values and we ``warm up" the training of the CNN for 25 epochs, using an exponential decay rate ($10^{-6}$) \citep{Huang_2016}  and then a staring learning rate of  0.001. Then the network was trained using an early stopping method, and for a maximum number of 300 epochs. The early stopping method is an effective method of preventing overfitting and consists in stopping the training if a monitored quantity does not improve for a fixed number (called \textit{patience}) of training epochs. The quantity that we monitored was the \textit{accuracy} of the classification of the validation sample. The best architecture was trained over 220 epochs with a parameter of patience equal to 20.
We implemented our code in the Keras framework \citep{Chollet_2015} on top of Theano \citep{Bastien_2012}.
Our architecture converges with a classification accuracy of 91\% on the validation sample. We further evaluated the performance of our classifier calculating the ROC (see section \ref{sec:figure_of_merit})  curve of the classifier, i.e. the True Positive Rate (TPR) against the False Positive Rate. As shown in Fig. \ref{Fig2_gamoclass}, we reached a TPR higher than the 90\% with a FPR < 8\%.

@@

We used a native CNN from Matlab\footnote{\texttt{https://www.mathworks.com/products/matlab.html}, \texttt{https://www.mathworks.com/help/nnet/convolutional-neural-networks.html}}, which has Convolutional 2D layers with 20 5x5 filters. This CNN can work either with one or three input images, representing greyscale and colour images.
We employed different strategies for the two samples available for the challenge, which involve combinations of the available bands running in one or more CNN, using or not the preprocessing, and combining the output with the aid of other machine learning methods.
In each case we used the simulations made available for the challenge both to train and to validate the results and we used the area under the ROC (see section \ref{sec:figure_of_merit}) to determine which combination of methods gives the best result.    
We selected 
%We evaluate the ROC curves for all configurations tested, selecting 
$90\%$ of the images, chosen randomly, for the training and $10\%$ to validate. We repeated the process $10$ times to avoid bias due to a specific choice of training/validation set and to define an uncertainty in our ROC.

@@

A more detailed description of each configuration tested is presented below:
\begin{enumerate}[label=\Roman*.]
\item Combination of bands $r$ and $i$ with the average between bands $u$ and $g$. Use one CNN for classification.
\item Creation of $1$ CNN for each band (total of $4$). The $4$ outputs are used as input to a SVM classifier which returns the final classification $p$. 
\item Combination into 4 different combinations of bands:  RGB $\rightarrow$ (\textit{u,g,r}), (\textit{u,g,i}), (\textit{u,r,i}) and (\textit{g,r,i}). One CNN for each combination of bands and then use of the output score as input to an SVM classifier.
%Mediatrix features used as input of an SVM classifier.
%
%IV) Use of wavelet decomposition for each band then use of these decomposed images in the same combinations of case III.
\item  Average of bands in different combinations RGB $\rightarrow$ (\textit{ug,r,i}), (\textit{u,gr,i}) and (\textit{u,g,ri}). The outputs of these 3 CNNs are inputs to a SVM classifier.
\item Use of CNN-activations (CNN feature maps) as inputs to a SVM classifier, using same combinations of bands of III. The output of each CNN is used as input to a SVM classifier.
\item Use of wiener filter and contrast adjustment on each band, then using the resulting images in the same architecture as in (III). 
\end{enumerate}

@@

\subsection{Figures of merit}
\label{sec:figure_of_merit}

@@

In addition to these considerations, the performance of a classifier is a function of many characteristics of the lens system.  It might be that one classifier is good at finding systems with large Einstein radii and incomplete arcs, but not as good at finding small complete Einstein rings that are blended with the light of the lens galaxy.  Also a lens may have a source that is too faint to be detected by any algorithm or is too far from the lens to be very distorted, but will be classified as a lens in the test dataset.    We do not impose a definitive arc/ring magnification, brightness or surface brightness limit for a system to be considered a lens because we 
want to include these ``barely lensed" objects to test the limits of the classifiers.  As we will see, if one restricts the objectives to detecting only lensed images with surface brightness above some threshold, for example, the "best" algorithm might change and the TPR will change.  For this reason we plot the AUROC, TPR$_0$ and TPR$_{10}$ as a function of several variables for all the entries.  This is done by removing all the lenses that do not exceed the threshold and then recalculating  these quantities, while the number of non-lenses remains the same.
